Pathogen Spillover - dissertation chapter

### 2019-04-07 d-sep_test.Rmd
??

---
output: html_document
---

@@

3. For each of the *k* pairs of variables (*Xi,Xj*), list the set of other variables, {*Z*} that are direct causes of either *Xi* or *Xj*. The pair of variables, (*Xi,Xj*), along with its conditioning set, {*Z*}, define an independence claim, (*Xi,Xj*) | {*Z*}, and the full set of the *k* independence claims defines the basis set, *Bu*.

4. For each element in this basis set, obtain the _**exact**_ probability, *Pk* that the pair (*Xi,Xj*) is statistically independent conditional on the variables *Z*. In other words, perform a regression model using an appropriate method.

@@

$$
C = -2\Sigma_{i=1}^k ln(P_i)
$$

@@

### Establish Basis Set
```{r get basis set, eval=TRUE}
library(ggm)
dag <- DAG(Live ~ Growth, Growth ~ Date, Date ~ DD, DD ~ Latitude)
isAcyclic(dag)# Logical test to check if the translated graph is acyclic, though a cyclic graph should return a warning, e.g.:
DAG(x~y, y~z, z~x)
Bu <- basiSet(dag)# Get the basis set as a list of vectors. The first two items of each vector are the names of the two non-adjacent nodes, followed by the names of the nodes in the conditioning set, which may be empty (though not in this instance).
```

Following the directions of the Shipley (2009) supplement I will fit separate mixed models for each independence claim in the basis set. 
```{r print basis set, echo=TRUE}
print(Bu)
```
Six claims equals six models. The first two variables listed in each item are claimed to be independent based on the following set of zero or more variables.

@@

# Print absolute correlations in either upper or lower panel; size proportianal to magnitude of the correlation
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(0, 1, 0, 1))
    r <- abs(cor(x, y, use="na.or.complete"))
    txt <- format(c(r, 0.123456789), digits = digits)[1]
    txt <- paste0(prefix, txt)
    if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
    text(0.5, 0.5, txt, cex = cex.cor * r)
}
```
```{r correlations & distributions}
#pairs(subset(data1, select = c(lat, Date, DD, Growth, Live)), diag.panel = panel.hist, lower.panel = panel.cor)
sub_data1 <- data1[ ,c(3,5:7,9)] #subset(data1, , select = c("lat","Date","DD","Growth","Live"))
#library(dplyr)
#sub_data1 <- data1 %>% select(lat, Date, DD, Growth, Live)

@@

The `Date` variable appears close to a normal distribution, so either the `lmer()` in the `lme4` package or the `lme()` function in the `nlme` package are appropriate. Shipley says that the `lme()` prints out degrees of freedom but `lmer()` does not. I am going to check to see if that is still true.
```{r independence claim 1 test}
fit1 <- lme(Date ~ DD + lat, data = data1, 
            random = ~1|site/tree, 
            na.action = na.omit)
fit1lme4 <- lmer(Date ~ DD + lat + (1|site) + (1|tree), data = data1,
                 na.action = na.omit)
summary(fit1)
summary(fit1lme4)
```
It is correct that `lmer()` does not provide degrees of freedom, and thus no probability estimate (p-value) for the coefficients. Aside from that it appears that the values estimate from both algorithms match up very well. Based on these results the null probability of observing a t-value at least as extreme for latitude (`lat`) is 0.9373, so pretty darn likely. That means that this independence claim is supported by the data.

Now to do the same for the other normally distributed dependent variables
```{r independence claim test 2 & 3}
# Independence of Growth and lat conditioned on Date
fit2 <- lme(Growth ~ Date + lat, data = data1, 
            random = ~1|site/tree, 
            na.action = na.omit)
# Independence of Growth and DD conditioned on Date and lat
fit3 <- lme(Growth ~ Date + lat + DD, data = data1, 
            random = ~1|site/tree, 
            na.action = na.omit)
```

>>>>

---
output:
  word_document: default
  html_document: default
---

@@

3. For each of the *k* pairs of variables (*X$_i$,X$_j$*), list the set of other variables, {*Z*} that are direct causes of either *X$_i$* or *X$_j$*. The pair of variables, (*X$_i$,X$_j$*), along with its conditioning set, {*Z*}, define an independence claim, (*X$_i$,X$_j$*) | {*Z*}, and the full set of the *k* independence claims defines the basis set, *B$_u$*.

4. For each element in this basis set, obtain the _**exact**_ probability, *P$_k$* that the pair (*X$_i$,X$_j$*) is statistically independent conditional on the variables *Z*. In other words, perform a regression model using an appropriate method.

@@

$$
C = -2\Sigma_{i=1}^k ln(P_i)
$$ 
6. Compare the value for *C* to a $\chi^2$ (chi-square) distribution with 2*k* degrees of freedom. If the P-value is greater than the *a priori* selected value, e.g. 0.05, then the data provides sufficient support for the model structure.

7. If the data provide sufficent support for the model structure, then each component may be assessed and coefficients interpreted.

@@

### Establish Basis Set

```{r get basis set, eval=TRUE}
library(ggm)
dag <- DAG(Live ~ Growth, Growth ~ Date, Date ~ DD, DD ~ Latitude)
isAcyclic(dag)# Logical test to check if the translated graph is acyclic, though a cyclic graph should return a warning, e.g.:
DAG(x~y, y~z, z~x)
Bu <- basiSet(dag)# Get the basis set as a list of vectors. The first two items of each vector are the names of the two non-adjacent nodes, followed by the names of the nodes in the conditioning set, which may be empty (though not in this instance).
```

Following the directions of the Shipley (2009) supplement I will fit separate mixed models for each independence claim in the basis set.

```{r print basis set, echo=TRUE}
print(Bu)
```

Six claims equals six models. The first two variables listed in each item are claimed to be independent based on the following set of zero or more variables.

@@

# Print absolute correlations in either upper or lower panel; size proportianal to magnitude of the correlation
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(0, 1, 0, 1))
    r <- abs(cor(x, y, use="na.or.complete"))
    txt <- format(c(r, 0.123456789), digits = digits)[1]
    txt <- paste0(prefix, txt)
    if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
    text(0.5, 0.5, txt, cex = cex.cor * r)
}
```

```{r correlations & distributions}
#pairs(subset(data1, select = c(lat, Date, DD, Growth, Live)), diag.panel = panel.hist, lower.panel = panel.cor)
sub_data1 <- data1[ ,c(3,5:7,9)] #subset(data1, , select = c("lat","Date","DD","Growth","Live"))
#library(dplyr)
#sub_data1 <- data1 %>% select(lat, Date, DD, Growth, Live)

@@

The `Date` variable appears close to a normal distribution, so either the `lmer()` in the `lme4` package or the `lme()` function in the `nlme` package are appropriate. Shipley says that the `lme()` prints out degrees of freedom but `lmer()` does not. I am going to check to see if that is still true.

```{r independence claim 1 test}
fit1 <- lme(Date ~ DD + lat, data = data1, 
            random = ~1|site/tree, 
            na.action = na.omit)
fit1lme4 <- lmer(Date ~ DD + lat + (1|site) + (1|tree), data = data1,
                 na.action = na.omit)
summary(fit1)
summary(fit1lme4)
```

It is correct that `lmer()` does not provide degrees of freedom, and thus no probability estimate (p-value) for the coefficients. Aside from that it appears that the values estimate from both algorithms match up very well. Based on these results the null probability of observing a t-value at least as extreme for latitude (`lat`) is 0.9373, so pretty darn likely. That means that this independence claim is supported by the data.

Now to do the same for the other normally distributed dependent variables

```{r independence claim test 2 & 3}
# Independence of Growth and lat conditioned on Date
fit2 <- lme(Growth ~ Date + lat, data = data1, 
            random = ~1|site/tree, 
            na.action = na.omit)
# Independence of Growth and DD conditioned on Date and lat
fit3 <- lme(Growth ~ Date + lat + DD, data = data1, 
            random = ~1|site/tree, 
            na.action = na.omit)
```
